abxD01 Paper

### 2015-06-07 Schubert_abxD01_mBio_2015.Rmd
PDS incorporated AMS and reviewers comments

```{r random_forest}
data <- read.table(file="data/process/random_forest.data")
n_full_features <- data[1,1]
rsq_full_features <- data[1,2] * 100

n_part_features <- data[2,1]
rsq_part_features <- data[2,2] * 100
```

@@

Perturbations to the gut microbiota result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*. Although *C. difficile* infection is commonly associated with antibiotic use, the precise alterations to the microbiota associated with this loss in function are unknown. We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which were then challenged with *C. difficile* spores. Across these treatments we observed that *C. difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner. Using relative abundance data, we built a machine learning regression model to predict the levels of *C. difficile* that were found 24 hours after challenging the perturbed communities. This model was able to explain `r round(rsq_full_features, 1)`% of the variation in the observed number of *C. difficile* per gram of feces. This model revealed important bacterial populations within the microbiota, which correlation analysis alone did not detect. Specifically, we observed that populations associated with the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes* were protective and populations associated with *Escherichia* and *Streptococcus* were associated with high levels of colonization. In addition, a population affiliated with *Akkermansia* indicated a strong context dependency on other members of the microbiota. Together, these results indicate that individual bacterial populations do not drive colonization resistance to *C. difficile*. Rather, multiple diverse assemblages act in concert to mediate colonization resistance.

@@

It is not completely understood how different perturbations to the gut
microbiota result in a loss of colonization resistance to *C.
difficile*. There is a clear need to better understand the ecology of
*C. difficile* and its interactions with members of the microbiota. In
mouse models of CDI, the unperturbed, untreated murine microbiome is
completely resistant to *C. difficile* colonization. It was previously
shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011
\#4;Theriot, 2011 \#7}, tigecycline {Bassis, 2014 \#8}, clindamycin
{Buffie, 2012 \#9}, or clindamycin in combination with a five antibiotic
cocktail {Chang, 2008 \#3} had decreased colonization resistance. These
studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a
bloom of *Lactobacillaceae* and *Enterobacteriaceae* are responsible for
the loss of colonization resistance. These results are largely supported
by human association studies {Schubert, 2014 \#11;Vincent, 2013 \#10}.
We previously observed significant differences between the gut microbiota of
hospitalized individuals with and without *C. difficile* and
non-hospitalized controls {Schubert, 2014 \#11}. In addition, fecal
microbiota transplants increase *Bacteroidetes* and decrease
*Proteobacteria* levels in recipients, resulting in a successful
restoration of colonization resistance in patients {Seekatz, 2014 \#12}.
Precisely how this occurs is not fully understood, but it emphasizes the
importance of the gut microbiota in colonization resistance against *C.
difficile*.

@@

**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to** ***C. difficile.*** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C. difficile* colonization (**Table 1**). Following the cessation of antibiotics, each treatment group was given one day of recovery prior to challenge with *C. difficile* spores. One day post challenge we enumerated the density of *C. difficile* in the animals' feces. We observed reproducibly high levels of *C. difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**). We observed variable levels of *C. difficile* colonization in mice treated with ampicillin. None of the mice that received ciprofloxacin were colonized. In addition to administering ciprofloxacin by oral gavage, we provided ciprofloxacin by intraperitoneal injection (10 mg/mL). For both approaches we provided one or two days of recovery. Regardless of the method, the resulting communities were resistant to *C. difficile* colonization. Only one of six mice receiving vancomycin was colonized with *C. difficile.* We suspected that this was due to residual vancomycin repressing *C. difficile* growth. In fact, two days post *C. difficile* challenge, *C. difficile* bloomed in these mice to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces. Furthermore, given two days of post-vancomycin recovery, there was no delay in *C. difficile* colonization to high levels, and on day one post challenge we observed a median of 3.0x10^7^ (interquartile range 2.6x10^7^–3.6x10^7^, N=4) CFU/g feces. These results suggest that although vancomycin is not absorbed by the gut tissue, the absence of *C. difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment. Overall, the various antibiotic perturbations provided varying levels of colonization by *C. difficile,* which suggested that the resulting communities varied in their composition.

To test this hypothesis, we sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C. difficile* challenge to identify populations within the microbiota that conferred colonization resistance. All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`). Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`). The community structures of mice receiving beta-lactams (i.e. cefoperazone and ampicillin) were not significantly different from each other (AMOVA, P=`r round(max(other_sig[other_comparison=="amp-cef"]), 2)`). These results indicate that perturbing the gut microbiota with antibiotics resulted in non-overlapping community structures that yielded significant variation in susceptibility to colonization when challenged with *C. difficile*.

@@

**Reduced perturbations result in altered levels of colonization.** Based on the *C. difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota. Consequently we expected a greater maintenance of resistance against *C. difficile* colonization in these titrated treatment groups. In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (*Figure S2*). These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e. cefoperazone), Gram-negative (i.e. streptomycin), and Gram-positive (i.e. vancomycin) bacteria. As expected in all mice receiving titrated doses of cefoperazone, *C. difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; *Figure 2*). Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs. We also observed increases in the relative abundances of OTUs associated with the family *Porphyromonadaceae* (OTU 5, 10, 11, 13, and 21; **Figure 2**). Reducing the dose of streptomycin significantly reduced the colonization levels (P\<`r round(max(strep.test$p.value, na.rm=T), 2)`; **Figure 2**). Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 5, 6, 10, and 11), *Alistipes* (OTU 12), and *Bacteroidales* (OTU 17). In addition to its anti-Gram-positive activity, vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of *C. difficile* colonization in only one mouse. We anticipated that lower doses might result in a community structure that would result in colonization. In fact, the 0.3 and 0.1 mg/mL doses of vancomycin resulted in similarly high levels of *C. difficile* colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`). Seven OTUs were differentially represented across the three doses of vancomycin. Surprisingly, even though the colonization levels of *C. difficile* did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose. Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 5, and 6) and one was affiliated with a member of the genus *Bacteroides* (OTU 1). Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level. These results suggest that the context in which specific members of the microbiota are found is important in determining the overall resistance to *C. difficile*. For example, the relationship between the *Bacteroides* (OTU 1) and *C. difficile* colonization was positive in streptomycin-treated mice and it was negative in cefoperazone-treated mice. In addition, cefoperazone and streptomycin-treated mice had high levels of *C. difficile* although the former had significantly higher levels of *Escherichia* (OTU 3), which were absent in the streptomycin-treated mice. Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance. In light of such results, resistance is likely a product of the overall composition of the community.

@@

**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we allowed the gut microbiota to recover for 24 hours before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored {Reeves, 2011 \#118; Theriot, 2014 \#956}. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**). Among the metronidazole-treated mice, those with extended recovery had a $`r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`$-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**). In addition, `r sum(metro_delay$CFU==0)` of the `r nrow(metro_delay)` mice given the longer recovery period had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**). Most notable among these was a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3). The relative abundance of this *Barnesiella* OTU increased with the delay, and the relative abundance of this *Escherichia* OTU decreased. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C. difficile* colonization (P=`r round(amp_test$p.value, 2)`). As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3). However untreated, fully resistant mice harbored significantly lower levels of *Barnesiella* (OTU 2). Rather, untreated mice had high levels of various *Porphyromonadaceae* OTUs (**Figure 1**). These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.

@@

**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C. difficile* across the three sets of experiments, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C. difficile* colonization 24 hours later (**Figure 4**). OTUs associated with providing resistance against *C. difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`). The *Porphyromonadaceae* (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C. difficile* colonization. Among the three *Proteobacteria* OTUs with a significant positive association with *C. difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`). By performing an OTU-based analysis we were able to observe intra-family and genus differences in association with *C. difficile* colonization. For example, the *Lachnospiraceae* have been associated with protection against *C. difficile*. Although we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C. difficile* colonization, `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU was associated with high levels of *C. difficile*. In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus* OTUs (family *Lactobacillaceae*), `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)` of which were associated with low levels of *C. difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels. The broad taxonomic representation of OTUs associated with low levels of *C. difficile* again suggests that a functionally diverse community is required to prevent the colonization *C. difficile*.

**The composition of the disturbed gut microbiota is predictive of** ***C. difficile*** **colonization levels.** These three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C. difficile* colonization, but in other contexts those OTUs had the opposite or no association. This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization. Correlation-based analyses cannot predict these types of context dependencies because they do not take into account the non-linearity and statistical interactions between populations. Therefore, we used a regression-based random forest machine learning algorithm to predict the level of *C. difficile* colonization observed in the three sets of experiments using the composition of the microbiota at the time of challenge as predictor variables. The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C. difficile* colonization levels (**Figures 5**). When we only included the top `r n_part_features` OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C. difficile* colonization levels. The OTUs that were ranked as being the most important in defining the random forest model further validated the observations from the correlation-based analysis (**Figures S4**). According to the random forest model, colonization resistance was associated with members of the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**). A loss in these populations, concurrently with a gain of *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), was associated with an increased susceptibility to infection (**Figure 6**). As we observed in the titration experiments, the relationship between *Akkermansia* (OTU 4) and *C. difficile* indicated that wide variation in the relative abundance of *Akkermansia* was associated with varying levels of *C. difficile*. There were varying abundances of *Akkermansia* in mice regardless of the level of *C. difficile* colonization. Finally, as indicated by the number of OTUs with relative abundances below the limit of detection, those mice could harbor varying levels of *C. difficile*. These observations bolster the hypothesis that colonization resistance is context dependent.

@@

There is clear need for more efficient therapies for treatment of *C. difficile* infections in humans aimed at restoration of the microbiota. Current first line treatments of CDI include regimens of either metronidazole or vancomycin, which further perturb the microbiota. As such, relapse rates for CDI are typically around 25-30% {Wilcox, 1998 \#19}. Interestingly, we observed that treatment with either antibiotic induced susceptibility to *C. difficile* in mice. This result has implications for understanding the causes of recurrent infections. Previous efforts to restore the microbiota and reestablish colonization resistance also support our findings. For instance, association of germ-free mice with a *Lachnospiraceae* isolate only reduced the level of *C. difficile* colonization by 10 to 100-fold {Reeves, 2012 \#18}. Using conventional mice, mixtures of bacteria rather than individual populations have been shown to restore colonization resistance and mediate clearance of *C. difficile* {Buffie, 2015 \#21;Lawley, 2012 \#20}. Fecal transplants, which represent a diverse collection of bacterial populations, have been highly effective in treating humans with recurrent *C. difficile* {Kassam, 2013 \#22;Seekatz, 2014 \#12;Weingarden, 2015 \#23}. By generating a diverse collection of communities that were challenged with *C. difficile*, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance. These would include members of the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes*. Several of these populations have been examined for their potential as a probiotic for preventing *C. difficile* infection. A 6-species cocktail, including isolates of *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, *Coriobacteriaceae*, *Staphylococcus*, and *Enterococcus*, successfully resolved CDI in mice {Lawley, 2012 \#20}. In humans, *Lactobacillus*-based probiotics have been co-administered with antibiotics to deter the onset of antibiotic-associated diarrhea (AAD) and *C. difficile* infection {Lawley, 2012 \#20}. A more diverse probiotic, which contained 33 bacterial species including *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, *Eubacteriaceae*, and *Lactobacillus* isolates, successfully restored colonization resistance in recurrent *C. difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 \#25}. Given this evidence, we feel confident that an effective probiotic community could be designed based on our findings to recover colonization resistance against *C. difficile*.

>>>>

```{r random_forest}
data <- read.table(file="data/process/random_forest.data")
n_full_features <- data[1,2]
rsq_full_features <- data[1,3] * 100

n_part_features <- data[2,2]
rsq_part_features <- data[2,3] * 100

binary_error <- data[3,3] * 100
```

@@

Perturbations to the gut microbiota can result in a loss of colonization resistance against gastrointestinal pathogens such as *Clostridium difficile*. Although *C. difficile* infection is commonly associated with antibiotic use, the precise alterations to the microbiota associated with this loss in function are unknown. We used a variety of antibiotic perturbations to generate a diverse array of gut microbiota structures, which were then challenged with *C. difficile* spores. Across these treatments we observed that *C. difficile* resistance was never attributable to a single organism, but rather it was the result of multiple microbiota members interacting in a context-dependent manner. Using relative abundance data, we built a machine learning regression model to predict the levels of *C. difficile* that were found 24 hours after challenging the perturbed communities. This model was able to explain `r round(rsq_full_features, 1)`% of the variation in the observed number of *C. difficile* per gram of feces. This model revealed important bacterial populations within the microbiota, which correlation analysis alone did not detect. Specifically, we observed that populations associated with the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes* were protective and populations associated with *Escherichia* and *Streptococcus* were associated with high levels of colonization. In addition, a population affiliated with *Akkermansia* indicated a strong context dependency on other members of the microbiota. Together, these results indicate that individual bacterial populations do not drive colonization resistance to *C. difficile*. Rather, multiple diverse assemblages act in concert to mediate colonization resistance.

@@

It is not completely understood how different perturbations to the gut
microbiota result in a loss of colonization resistance to *C.
difficile*. There is a clear need to better understand the ecology of
*C. difficile* and its interactions with members of the microbiota. In
mouse models of CDI, the unperturbed, untreated murine microbiome is
completely resistant to *C. difficile* colonization. It was previously
shown that C57Bl/6 mice treated with cefoperazone {Reeves, 2011
\#4;Theriot, 2011 \#7}, tigecycline {Bassis, 2014 \#8}, clindamycin
{Buffie, 2012 \#9}, or clindamycin in combination with a five antibiotic
cocktail {Chang, 2008 \#3} had decreased colonization resistance. These
studies suggest that a loss of *Lachnospiraceae* and *Barnesiella* and a
bloom of *Lactobacillaceae* and *Enterobacteriaceae* are responsible for
the loss of colonization resistance. These results are largely supported
by human association studies {Schubert, 2014 \#11;Vincent, 2013 \#10}.
We previously observed significant differences between the gut microbiota of
hospitalized individuals with and without *C. difficile* and
non-hospitalized controls {Schubert, 2014 \#11}. In addition, fecal
microbiota transplants have been shown to increase the relative abundance of *Bacteroidetes* and decrease the relative abundance of *Proteobacteria* and result in a successful
restoration of colonization resistance in patients {Seekatz, 2014 \#12}.
The mechanisms involved in restoring colonization resistance is not fully understood, but it emphasizes the importance of the gut microbiota in protecting against *C.
difficile*.

@@

**Antibiotics differentially alter the structure of the microbiota and their colonization resistance to** ***C. difficile.*** We selected a panel of seven antibiotics from six classes with the goal of differentially altering the microbiota and assessing their resistance to *C. difficile* colonization (**Table 1**). Following the cessation of antibiotics, each treatment group was given one day of recovery prior to challenge with *C. difficile* spores. One day post challenge we enumerated the density of *C. difficile* in the animals' feces. We observed reproducibly high levels of *C. difficile* colonization in mice treated with cefoperazone, metronidazole, and streptomycin (**Figures 1 and S1**). We observed the most variance in levels of *C. difficile* colonization in mice treated with ampicillin. None of the mice that received ciprofloxacin were colonized. In addition to administering ciprofloxacin by oral gavage, we provided ciprofloxacin by intraperitoneal injection (10 mg/mL). For both approaches we provided one or two days of recovery from the antibiotic treatment. Regardless of the method, the resulting communities were resistant to *C. difficile* colonization. Only one of six mice receiving vancomycin was colonized with *C. difficile.* We suspected that this was due to residual vancomycin repressing *C. difficile* growth. In fact, two days post *C. difficile* challenge, *C. difficile* bloomed in this treatment group to a median of 9.1x10^7^ (interquartile range 7.6x10^7^–1.1x10^8^) CFU/g feces. Furthermore, given two days of post-vancomycin recovery, there was no delay in *C. difficile* colonization to high levels, and on day one post challenge we observed a median of 3.0x10^7^ (interquartile range 2.6x10^7^–3.6x10^7^, N=4) CFU/g feces. These results suggest that although vancomycin is not absorbed by the gut tissue, the absence of *C. difficile* in the remaining five vancomycin-treated mice may have been due to residual antibiotics lingering in the environment. Overall, the various antibiotic perturbations provided varying levels of colonization by *C. difficile,* which suggested that the resulting communities varied in their composition.

To test this hypothesis, we sequenced the 16S rRNA genes from the fecal communities of treated and untreated mice prior to *C. difficile* challenge to identify populations within the microbiota that conferred colonization resistance. All of the antibiotic treatments, except for the ciprofloxacin-treated mice (AMOVA, P=`r round(max(control_sig[control_comparison=="cipro-control"]), 2)`), resulted in distinct and reproducible changes to the structure of the microbiota relative to the untreated animals (AMOVA, P<`r ifelse(max(control_sig[control_comparison!="cipro-control"]) < 1e-3, "0.001", "NA")`). The similarity in the structure of the microbiota in ciprofloxacin-treated and untreated mice suggests that a higher dose of ciprofloxacin may have been necessary to significantly perturb the microbiota to allow *C. difficile* to overcome colonization resistance. Comparisons of the microbiota between antibiotic classes indicated that their structures were significantly different from each other (AMOVA, P<`r round(max(other_sig[other_comparison!="amp-cef"]), 2)`). The community structures of mice receiving beta-lactams (i.e. cefoperazone and ampicillin) were not significantly different from each other (AMOVA, P=`r round(max(other_sig[other_comparison=="amp-cef"]), 2)`). These results indicate that perturbing the gut microbiota with antibiotics resulted in non-overlapping community structures that yielded significant variation in susceptibility to colonization when challenged with *C. difficile*.

@@

**Reduced perturbations result in altered levels of colonization.** Based on the *C. difficile* colonization levels in our seven antibiotic treatments, we hypothesized that titrating the dose of antibiotics that the mice received would result in smaller perturbations to the microbiota. Consequently we expected a greater maintenance of resistance against *C. difficile* colonization in these titrated treatment groups. In addition to the previous treatments, we treated mice with lower concentrations of cefoperazone, streptomycin, and vancomycin  (*Figure S2*). These antibiotics were selected because they are thought to target a broad spectrum of bacteria (i.e. cefoperazone), Gram-negative (i.e. streptomycin), and Gram-positive (i.e. vancomycin) bacteria. As expected in all mice receiving titrated doses of cefoperazone, *C. difficile* colonization levels decreased significantly (P\<`r round(max(cef.test$p.value, na.rm=T), 2)`; *Figure 2*). Titrating the dose of cefoperazone in the animals' drinking water resulted in significant decreases in the relative abundance of an OTU associated with the genus *Escherichia* (OTU 3) and a number of rare OTUs. We also observed increases in the relative abundances of OTUs associated with the family *Porphyromonadaceae* (OTU 5, 10, 11, 13, and 21; **Figure 2**). Reducing the dose of streptomycin significantly reduced the colonization levels (P\<`r round(max(strep.test$p.value, na.rm=T), 2)`; **Figure 2**). Titrating the dose of streptomycin in the drinking water resulted in significant changes in the relative abundance of OTUs associated with the *Porphyromonadaceae* (OTUs 2, 5, 6, 10, and 11), *Alistipes* (OTU 12), and *Bacteroidales* (OTU 17). In addition to its anti-Gram-positive activity, vancomycin was also selected because although the community was quite different from untreated mice, we observed high levels of *C. difficile* colonization in only one mouse. We anticipated that lower doses might result in a community structure that would result in colonization. In fact, the 0.3 and 0.1 mg/mL doses of vancomycin resulted in similarly high levels of *C. difficile* colonization (P=`r round(vanc.test$p.value["0.3","0.1"], 2)`). Seven OTUs were differentially represented across the three doses of vancomycin. Surprisingly, even though the colonization levels of *C. difficile* did not significantly differ between the mice receiving 0.1 and 0.3 mg/mL of vancomycin in their drinking water, four of the OTUs that had significantly different relative abundances were only found in the lower dose. Three of these were affiliated with members of the *Porphyromonadaceae* (OTUs 2, 5, and 6) and one was affiliated with a member of the genus *Bacteroides* (OTU 1). Two OTUs affiliated with the *Akkermansia* (OTU 6) and *Lactobacillus* (OTU 8) genera increased with increasing dose and a third OTU affiliated with *Escherichia* (OTU 4) had a mixed response to the dose level. These results suggest that the context in which specific members of the microbiota are found is important in determining the overall resistance to *C. difficile*. For example, the relationship between the *Bacteroides*-affiliated OTU (OTU 1) and *C. difficile* colonization was positive in streptomycin-treated mice and it was negative in cefoperazone-treated mice. In addition, cefoperazone and streptomycin-treated mice had high levels of *C. difficile* although the former had significantly higher levels of an *Escherichia*-affiliated OTU (OTU 3), which were absent in the streptomycin-treated mice. Together, these results suggest that individual populations were not sufficient to consistently predict colonization resistance. In light of such results, resistance is likely a product of the overall composition of the community.

@@

**Allowing recovery of the microbiota restores colonization resistance.** In the experiments we have described thus far, we allowed the gut microbiota to recover for 24 hours before challenging them with *C. difficile*. Several studies have demonstrated that perturbed communities can return to a "healthy" state in which resistance to *C. difficile* is restored {Reeves, 2011 \#118; Theriot, 2014 \#956}. To test the effect of recovery on colonization and gain greater insights into the populations that confer colonization resistance, we allowed the microbiota of the mice that received the full metronidazole and ampicillin treatment to recover for an additional five days (**Figure S3**). Among the metronidazole-treated mice, those with extended recovery had a $`r format(metro_med_cfu[2]/metro_med_cfu[1], sci=T)`$-fold reduction in colonization (P<`r ifelse(metro_test$p.value<0.001, "0.001", "NA")`; **Figure 3**). In addition, `r sum(metro_delay$CFU==0)` of the `r nrow(metro_delay)` mice given the longer recovery period had no detectable *C. difficile* 24 hours after challenge. We detected six OTUs that were differentially represented in the two sets of metronidazole-treated mice (**Figure 3**). Most notable among these were two OTUs that affiliated with a member of the *Barnesiella* (OTU 2) and the *Escherichia* (OTU 3). The relative abundance of this *Barnesiella*-affiliated OTU increased with the delay, and the relative abundance of this *Escherichia*-affiliated OTU decreased. Similar to the metronidazole-treated mice, the ampicillin-treated mice that were allowed to recover an additional five days before challenge had a significant decrease in *C. difficile* colonization (P=`r round(amp_test$p.value, 2)`). As before, we observed a similar increase and decrease in relative abundances for *Barnesiella* (OTU 2) and *Escherichia* (OTU 3)-affiliated OTUs. However untreated, fully resistant mice harbored significantly lower levels of the *Barnesiella*-affiliated OTU (OTU 2). Rather, untreated mice had high levels of various *Porphyromonadaceae*-affiliated OTUs (**Figure 1**). These findings further confirm the context-dependency of colonization resistance suggested by the results of our titration experiments.

@@

**Correlation analysis reveals potentially protective bacteria.** To identify bacterial taxa that could be associated with resistance or susceptibility to *C. difficile* across the three sets of experiments, we measured the correlation between the relative abundance of each OTU on the day of inoculation with the level of *C. difficile* colonization 24 hours later (**Figure 4**). OTUs associated with providing resistance against *C. difficile* (N=`r sum(corr_data$sig_corrs < 0)`) outnumbered those with associated with susceptibility (N=`r sum(corr_data$sig_corrs > 0)`). The *Porphyromonadaceae*-affiliated OTUs (ρ<sub>average</sub>=`r ave_corr_family["Porphyromonadaceae"]`, N=`r N_family["Porphyromonadaceae"]` OTUs) were consistently associated with low levels of *C. difficile* colonization. Among the three *Proteobacteria*-affiliated OTUs with a significant positive association with *C. difficile* colonization, the strongest was affiliated with the *Escherichia* (OTU 3; ρ=`r round(corr_data[families=="Enterobacteriaceae","sig_corrs"], 2)`). By performing an OTU-based analysis we were able to observe intra-family and genus differences in association with *C. difficile* colonization. For example, the *Lachnospiraceae* have been associated with protection against *C. difficile*. Although within the *Lachnospiraceae* familiy we observed `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] < 0)` OTUs that were associated with low levels of *C. difficile* colonization, `r sum(corr_data[families=="Lachnospiraceae","sig_corrs"] > 0)` OTU was associated with high levels of *C. difficile*. In addition we observed `r nrow(corr_data[families=="Lactobacillaceae",])` significantly correlated *Lactobacillus*-affiliated OTUs (family *Lactobacillaceae*, `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] < 0)`) of which were associated with low levels of *C. difficile* and `r sum(corr_data[families=="Lactobacillaceae","sig_corrs"] > 0)` was associated with high levels. The broad taxonomic representation of OTUs associated with low levels of *C. difficile* again suggests that a diverse community may be advantageous in preventing *C. difficile* colonization.

**The composition of the disturbed gut microbiota is predictive of** ***C. difficile*** **colonization levels.** These three sets of experiments demonstrated that in certain contexts individual OTUs could be associated with *C. difficile* colonization, but in other contexts those OTUs had the opposite or no association. This suggests that colonization is a phenotype that is driven by multiple populations that act independently and possibly in concert to resist colonization. Correlation-based analyses cannot predict these types of context dependencies because they do not take into account the non-linearity and statistical interactions between populations. Therefore, we used a regression-based random forest machine learning algorithm to predict the level of *C. difficile* colonization observed in the three sets of experiments using the composition of the microbiota at the time of challenge as predictor variables. The model explained `r round(rsq_full_features, 1)`% of the variation in the observed *C. difficile* colonization levels (**Figures 5**). When we only included the top `r n_part_features` OTUs based on the percent increase in the mean squared error when each OTU was removed, the resulting model explained `r round(rsq_part_features, 1)`% of the variation in the observed *C. difficile* colonization levels. The OTUs that were ranked as being the most important in defining the random forest model further validated the observations from the correlation-based analysis (**Figures S4**). According to the random forest model, colonization resistance was associated with OTUs that affiliated with the *Porphyromonadaceae* (OTU 15, 10, 6, 18, and 11), *Lachnospiraceae* (OTU 25), *Lactobacillus* (OTU 23), *Alistipes* (OTU 12), and *Turicibacter* (OTU 9; **Figure 6**). A loss in these populations, concurrently with a gain in OTUs affiliated with the *Escherichia* (OTU 3) or *Streptococcus* (OTU 90), was associated with an increased susceptibility to infection (**Figure 6**). As we observed in the titration experiments, the relationship between an *Akkermansia*-affiliated OTU (OTU 4) and *C. difficile* indicated that wide variation in the relative abundance of *Akkermansia* was associated with varying levels of *C. difficile*. There were varying abundances of the *Akkermansia*-affiliated OTU in mice regardless of the level of *C. difficile* colonization. Finally, as indicated by the number of OTUs with relative abundances below the limit of detection, those mice could harbor varying levels of *C. difficile*. These observations bolster the hypothesis that colonization resistance is context dependent.

@@

There is clear need for more efficient therapies for treatment of *C. difficile* infections in humans aimed at restoration of the microbiota. Current first line treatments of CDI include regimens of either metronidazole or vancomycin, which further perturb the microbiota. As such, relapse rates for CDI are typically around 25-30% {Wilcox, 1998 \#19}. Interestingly, we observed that treatment with either antibiotic induced susceptibility to *C. difficile* in mice. This result has implications for understanding the causes of recurrent infections. Previous efforts to restore the microbiota and reestablish colonization resistance also support our findings. For instance, association of germ-free mice with a *Lachnospiraceae* isolate only reduced the level of *C. difficile* colonization by 10 to 100-fold {Reeves, 2012 \#18}. Using conventional mice, mixtures of bacteria rather than individual populations have been shown to restore colonization resistance and mediate clearance of *C. difficile* {Buffie, 2015 \#21;Lawley, 2012 \#20}. Fecal transplants, which represent a diverse collection of bacterial populations, have been highly effective in treating humans with recurrent *C. difficile* {Kassam, 2013 \#22;Seekatz, 2014 \#12;Weingarden, 2015 \#23}. By generating a diverse collection of communities that were challenged with *C. difficile*, we have identified a subset of populations using random forest modeling that could be used as a probiotic cocktail to provide colonization resistance. These would include members of the *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, and *Alistipes*. Several of these populations have been examined for their potential as a probiotic for preventing *C. difficile* infection. A 6-species cocktail, including isolates of *Porphyromonadaceae*, *Lachnospiraceae*, *Lactobacillus*, *Coriobacteriaceae*, *Staphylococcus*, and *Enterococcus*, successfully resolved CDI in mice {Lawley, 2012 \#20}. In humans, *Lactobacillus*-based probiotics have been co-administered with antibiotics to deter the onset of antibiotic-associated diarrhea (AAD) and *C. difficile* infection {Lawley, 2012 \#20}. A more diverse probiotic, which contained 33 bacterial species including *Porphyromonadaceae*, *Lachnospiraceae*, *Ruminococcaceae*, *Eubacteriaceae*, and *Lactobacillus* isolates, successfully restored colonization resistance in recurrent *C. difficile* infection and eliminated diarrhea up to 6 months post treatment {Petrof, 2013 \#25}. Given this evidence, we feel confident that an effective probiotic mixture could be designed based on our findings to recover colonization resistance against *C. difficile*. Moreover, this line of study will be useful towards the development of personalized treatments based on an individual's specific gut microbiota, which may be a more efficient strategy for preventing and treating CDI. Further examination of the bacterial populations identified in this study is necessary to identify causal relationships and assess the specific mechanisms of colonization resistance. Such research will further advance the development of protocols to prevent and treat CDI.
